Non-word repetition in Yélî Dnye

### 2021-02-12 manuscript.Rmd
submitted

@@

keywords          : "phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness"
wordcount         : "~11,500 words"

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001non-word], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas, three via direct results.

@@

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006non-word], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006non-word]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021non-word]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014non-word; @vance2005speech]. Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014non-word; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological]. We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014non-word], with the entire test inventory of 40 final items distributed across children.

@@

In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006non-word; but note  @cristia2020infant is an exception]. Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency. In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014non-word; @meir2017independent]  on NWR scores. We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r~.1, which is small. Similarly, NWR scores may not vary greatly with participant gender according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006non-word; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014non-word and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).

>>>>

author_note: >
       Both authors contributed to study funding, design, data collection, annotation, analyses, writing.

@@

keywords          : "phonology, non-word repetition, Papuan, non-industrial, non-urban, comparative, typology, markedness, literacy"
wordcount         : "~11,500 words"

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas, three via direct results.

@@

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

@@

The third research area we contribute data to relates to the possibility that individual variation in NWR scores is structured. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education. Based on previous work, we looked at potential increases with age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Prior research typically finds no significant differences as a function of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio] or child gender [@chiat2007preschool]. Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological]. We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-30 items: @jabere2018xperiment; @kalnak2014nonword], with the entire test inventory of 40 final items distributed across children.

@@

In broad terms, we expected that NWR scores would increase with participant age, as this is the pattern observed in several of the studies in Figure \@ref(fig:fig-prevlit) [English @vance2005speech; Italian @piazzalunga2019articulatory; Cantonese @stokes2006nonword; but note  @cristia2020infant is an exception]. Indeed, age was significantly correlated with NWR score and also showed up as a significant predictor of NWR score when included as a control factor in the analyses of both item length and average segmental frequency. In brief, our results underscore the idea that phonological development continues well past the first few years of life, extending into middle childhood and perhaps later [@hazan2000development].

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores. We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r~.1, which is small. Similarly, NWR scores may not vary greatly with participant gender according to previous work [@chiat2007preschool], and for that as well we find effects of about that size.

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye from the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure \@ref(fig:fig-prevlit-overall).
